Cross Validation Approaches for Penalized Cox Regression

### 2018-11-12 main.tex
oracle MSE and CV-AUC

\par We first measure the distance between the fitted model and the true model via mean squared error of the coefficients. Suppose $\hat{\beta}(\lambda)$ is the set of coefficients estimated by a cox model with LASSO penalty at $\lambda$. Mean squared error $MSE$ is defined to be 
\begin{equation}
E(\hat{\beta}(\lambda) - \beta)^2
\end{equation}
For each generated data set, the $\lambda$ that has the minimal MSE is chosen as the optimal $\lambda$, which we denote by $\lambda^{opt}$. Then the models selected by the cross validation methods with $\lambda^{cv}$ would be compared to the model that corresponds to $\lambda^{opt}$. We would take the ratio between $MSE$ of $\hat{\beta}(\lambda^{cv})$ and $MSE$ of $\hat{\beta}(\lambda^{optimal})$ for each cross-validation selected $\lambda^{cv}$ in each generated data set. The results would be averaged over N replications.

@@

\caption{\label{tab:} Simulation Studies that Compare All Four Cross-Validation Methods Varying Dimension and Signal }
\centering
\begin{tabular}[t]{cccccccccc}
\toprule
 n& p& $\#$ Non-Zero & Signal & Method & $\lambda$ & MSE Ratio &Brier Score & Kullback Leibler & C Index\\
\midrule
150 & 50 & 5 & Weak & V$\&$ VH  & 0.095 (0.028) & 1.323 (0.443) & 0.242 (0.013) & 0.678 (0.029) & 0.617 (0.034) \\
      &    &   &     & Standard  & 0.087 (0.030) & 1.323 (0.431) & 0.242 (0.013) & 0.678 (0.029) & 0.618 (0.032) \\
      &    &   &     & LinearPred  & 0.088 (0.025) & 1.285 (0.419) & 0.241 (0.013) & 0.677 (0.028) & 0.620 (0.029) \\
      &    &   &     & DevResid  & 0.132 (0.032) & 1.697 (0.716) & 0.248 (0.014) & 0.689 (0.030) & 0.601 (0.044) \\
\addlinespace
150 & 50 & 5 & Strong & V$\&$VH  & 0.070 (0.011) & 1.373 (0.362) & 0.177 (0.012) & 0.527 (0.030) & 0.750 (0.014) \\
    &    &   &     & Standard  & 0.066 (0.012) & 1.313 (0.358) & 0.176 (0.012) & 0.526 (0.029) & 0.749 (0.014) \\
    &    &   &     & LinearPred  & 0.067 (0.011) & 1.319 (0.335) & 0.176 (0.012) & 0.526 (0.030) & 0.750 (0.014) \\
    &    &   &     & DevResid  & 0.110 (0.012) & 2.569 (0.826) & 0.187 (0.014) & 0.554 (0.033) & 0.750 (0.018) \\
\addlinespace
300 & 10000 & 20 & Weak & V$\&$VH  & 0.139 (0.021) & 1.353 (0.192) & 0.244 (0.015) & 0.681 (0.032) & 0.612 (0.051) \\
      &    &   &     & Standard  & 0.105 (0.027) & 1.206 (0.150) & 0.228 (0.021) & 0.646 (0.045) & 0.639 (0.050) \\
      &    &   &     & LinearPred  & 0.107 (0.026) & 1.218 (0.159) & 0.229 (0.021) & 0.648 (0.045) & 0.638 (0.049) \\
      &    &   &     & DevResid  & 0.139 (0.027) & 1.341 (0.198) & 0.243 (0.018) & 0.678 (0.038) & 0.610 (0.058) \\
\addlinespace
500 & 10000 & 20 & Strong & V$\&$VH  & 0.072 (0.005) & 1.837 (0.184) & 0.162 (0.015) & 0.495 (0.036) & 0.784 (0.021) \\
      &    &   &     & Standard  & 0.056 (0.005) & 1.368 (0.146) & 0.153 (0.011) & 0.465 (0.029) & 0.782 (0.017) \\
      &    &   &     & LinearPred  & 0.057 (0.005) & 1.410 (0.150) & 0.153 (0.011) & 0.466 (0.029) & 0.783 (0.017) \\
      &    &   &     & DevResid  & 0.076 (0.005) & 1.985 (0.227) & 0.167 (0.015) & 0.508 (0.035) & 0.782 (0.022) \\
\bottomrule
\end{tabular}
\end{table}

@@

@@

\par While the Deviance Residual approach has a reasonable performance in scenarios where dimension is high, it tends to act more conservatively than all the other three approaches in general. As is noted by Therneau et al, the sums of squared martingale residuals and deviance residuals do not necessary reflect how good the model fit is \citep{Therneau2000modeling}. The difficulty of estimating the baseline hazard seems to play an important role. A series of simulations were thus designed to study how baseline hazard estimation affects the deviance residuals' performance as a measurement for model fit.

\par In the following simulation studies, the defined optimal $\lambda$ and the $\lambda$ selected by the Linear Predictor approach were used as benchmarks. Various sample sizes and dimensions were varied. The survival outcomes were generated both from exponential distribution and Weibull distribution. Six different approaches were compared in terms of baseline hazard estimation and their impact on sums of squared deviance residuals. Two non-parametric approaches, Kalbfleisch and Prentices' estimator and Breslow's estimator, were used first for estimating the cumulative hazard baseline \citep{Kalbfleisch2011} \citep{breslow1972}.  As is presented in column 3 and column 5 in Table 2, the sum of squared deviance residuals computed based on both approaches are all very conservative compared to the optimal $\lambda$. Their performance is extremely bad when n is large and p is small.

\par Since both of the non-parameteric estimations seemed to perform well at early time points but poorly later, a weighted approach was then applied to emphasize earlier time points. Weights are assigned proportional to number of subjects at risk, making earlier time points weighted more than later time points. Results of these two approahces are presented in column 4 and 6 in Table 2. The attempt to improve the non-parameteric estimations via weighting does not seem to improve the selection too much.

\par In the next step,  the non-parametric baseline hazard estimation was replaced by the exact baseline hazard that was used to generate the outcome. Results were listed in the seventh column in Table 2. When the baseline was known, the performance of sum of squared deviance residuals actually performed really close to the optimal $\lambda$. This simulation suggests that when the baseline hazard was accurately estimated, then the sum of squared deviance residuals gave really good measures for the model fit.

\par In the last step, we replaced the non-parametric baseline hazard estimation with a parametric approach, where we assume the baseline hazard to follow an exponential distribution. We derived the estimation for baseline hazard by maximum likelihood approach conditioned on covariates: $\hat{\lambda}_{0} = \frac{\sum d_{i}}{\sum t_{i}e^{\eta_{i}}}$. The results are listed in the eighth column in Table 2. When this estimation was used for data that were generated from exponential distribution, its performance was as well as when the exact baseline was specified. When it was fitted to data that were generated from the Weibull distribution, that is when the model is misspecified, then its performance was even worse than the non-parametric approaches.

\par Hence, getting accurate estimation of the baseline hazard is crucial to whether or not the sum of squared deviance residuals can be a good measurement for model fit.

\begin{table}[h]
	\small
	\caption{Average of Selected $\lambda$ with Different Baseline Estimation for Deviance Residuals}
	\centering
	\begin{tabular}{lllllllll}
	\hline
	\\[-0.75em]
    & Optimal & Linear Pred & KP & KP & Breslow & Breslow & Exact & Exp\\ 
    &  & & & weighted & & weighted & & \\ \cline{2 - 9} 
	\textbf{True Baseline: $Exponential(1)$} &&&&&&&&\\
	n = 100, p = 10, non-zero $\beta$: 2 & 0.0461 & 0.0665 & 0.1986 & 0.1970 & 0.1908 & 0.1975 & 0.0540 & 0.0539 \\
	n = 100, p = 100, non-zero $\beta$: 20 &0.0555 & 0.1841 & 0.2359 & 0.2201 & 0.2146 & 0.2244 & 0.0805 & 0.0985 \\
	n = 100, p = 1000, non-zero $\beta$: 20 & 0.0268 & 0.0505 & 0.0948 & 0.0798 & 0.0867 & 0.0809 & 0.0288 & 0.0270 \\
	\textbf{True Baseline:} $Weibull$ &&&&&&&& \\
 	n = 100, p = 10, non-zero $\beta$: 2 & 0.0476 & 0.0710 & 0.2004 & 0.1969 & 0.1907 & 0.1907 & N/A & 0.2330 \\ 
	\hline
\end{tabular}
\end{table}

>>>>

\par We first measure the distance between the fitted model and the true model via mean squared error of the coefficients. Suppose $\hat{\beta}(\lambda)$ is the set of coefficients estimated by a cox model with LASSO penalty at $\lambda$. Mean squared error $MSE$ is defined to be 
\begin{equation}
E(\hat{\beta}(\lambda) - \beta)^2
\end{equation}
For each generated data set, an oracle model would be fitted using Cox regression only upon variables that are set as non-zero in the data generating process. The resulted estimates would be denoted by $\hat{\beta}^{oracle}$. Then the models selected by the cross validation methods with $\lambda^{cv}$ would be compared to the oracle model via 
\begin{equation}
log\left\{ \frac{MSE(\hat{\beta}(\lambda^{cv}))}{MSE(\hat{\beta}^{oracle})} \right\}
\end{equation}
for each cross-validation selected $\lambda^{cv}$ in each generated data set. The results would be averaged over N replications.

%For each generated data set, the $\lambda$ that has the minimal MSE is chosen as the optimal $\lambda$, which we denote by $\lambda^{opt}$. Then the models selected by the cross validation methods with $\lambda^{cv}$ would be compared to the model that corresponds to $\lambda^{opt}$. We would take the ratio between $MSE$ of $\hat{\beta}(\lambda^{cv})$ and $MSE$ of $\hat{\beta}(\lambda^{optimal})$ for each cross-validation selected $\lambda^{cv}$ in each generated data set. The results would be averaged over N replications.

@@

\caption{\label{tab:} Simulation Studies that Compare All Four Cross-Validation Methods Varying Dimension and Signal }
\centering
\begin{tabular}[t]{cccccccccc}
\toprule
 n& p& $\#$ Non-Zero & Signal & Method & $\lambda$ & log(MSE Ratio) &Brier Score & Kullback Leibler & C Index\\
\midrule
150 & 50 & 5 & 0.3 & V VH  & 0.095 (0.028) & 1.457 (0.954) & 0.242 (0.013) & 0.678 (0.029) & 0.617 (0.034) \\
    &    &   &     & Standard  & 0.087 (0.030) & 1.459 (0.922) & 0.242 (0.013) & 0.678 (0.029) & 0.618 (0.032) \\
    &    &   &     & LinearPred  & 0.088 (0.025) & 1.433 (0.927) & 0.241 (0.013) & 0.677 (0.028) & 0.620 (0.029) \\
    &    &   &     & DevResid  & 0.132 (0.032) & 1.676 (1.019) & 0.248 (0.014) & 0.689 (0.030) & 0.601 (0.044) \\
\addlinespace
150 & 50 & 5 & 0.6 & V VH  & 0.070 (0.011) & 1.733 (0.922) & 0.177 (0.012) & 0.527 (0.030) & 0.750 (0.014) \\
    &    &   &     & Standard  & 0.066 (0.012) & 1.688 (0.909) & 0.176 (0.012) & 0.526 (0.029) & 0.749 (0.014) \\
    &    &   &     & LinearPred  & 0.067 (0.011) & 1.695 (0.925) & 0.176 (0.012) & 0.526 (0.030) & 0.750 (0.014) \\
    &    &   &     & DevResid  & 0.110 (0.012) & 2.339 (0.868) & 0.187 (0.014) & 0.554 (0.033) & 0.750 (0.018) \\
\addlinespace
300 & 10000 & 20 & low & V VH  & 0.139 (0.021) & 3.463 (0.438) & 0.244 (0.015) & 0.681 (0.032) & 0.612 (0.051) \\
    &    &   &     & Standard  & 0.105 (0.027) & 3.350 (0.458) & 0.228 (0.021) & 0.646 (0.045) & 0.639 (0.05) \\
    &    &   &     & LinearPred  & 0.107 (0.026) & 3.360 (0.462) & 0.229 (0.021) & 0.648 (0.045) & 0.638 (0.049) \\
    &    &   &     & DevResid  & 0.139 (0.027) & 3.454 (0.447) & 0.243 (0.018) & 0.678 (0.038) & 0.610 (0.058) \\
\addlinespace
500 & 10000 & 20 & strong & V VH  & 0.072 (0.005) & 3.469 (0.480) & 0.162 (0.015) & 0.495 (0.036) & 0.784 (0.021) \\
    &    &   &     & Standard  & 0.056 (0.005) & 3.174 (0.504) & 0.153 (0.011) & 0.465 (0.029) & 0.782 (0.017) \\
    &    &   &     & LinearPred  & 0.057 (0.005) & 3.204 (0.498) & 0.153 (0.011) & 0.466 (0.029) & 0.783 (0.017) \\
    &    &   &     & DevResid  & 0.076 (0.005) & 3.545 (0.467) & 0.167 (0.015) & 0.508 (0.035) & 0.782 (0.022) \\
\bottomrule
\end{tabular}
\end{table}

@@

 \subsection {Cross-Validated C Index}

\begin{figure}[h]
    \centering
		\includegraphics[height= 15cm ]{./figures/figure_auc.png}
    \caption{Cross-Validated AUC}
\end{figure}

@@

%\par While the Deviance Residual approach has a reasonable performance in scenarios where dimension is high, it tends to act more conservatively than all the other three approaches in general. As is noted by Therneau et al, the sums of squared martingale residuals and deviance residuals do not necessary reflect how good the model fit is \citep{Therneau2000modeling}. The difficulty of estimating the baseline hazard seems to play an important role. A series of simulations were thus designed to study how baseline hazard estimation affects the deviance residuals' performance as a measurement for model fit.

%\par In the following simulation studies, the defined optimal $\lambda$ and the $\lambda$ selected by the Linear Predictor approach were used as benchmarks. Various sample sizes and dimensions were varied. The survival outcomes were generated both from exponential distribution and Weibull distribution. Six different approaches were compared in terms of baseline hazard estimation and their impact on sums of squared deviance residuals. Two non-parametric approaches, Kalbfleisch and Prentices' estimator and Breslow's estimator, were used first for estimating the cumulative hazard baseline \citep{Kalbfleisch2011} \citep{breslow1972}.  As is presented in column 3 and column 5 in Table 2, the sum of squared deviance residuals computed based on both approaches are all very conservative compared to the optimal $\lambda$. Their performance is extremely bad when n is large and p is small.

%\par Since both of the non-parameteric estimations seemed to perform well at early time points but poorly later, a weighted approach was then applied to emphasize earlier time points. Weights are assigned proportional to number of subjects at risk, making earlier time points weighted more than later time points. Results of these two approahces are presented in column 4 and 6 in Table 2. The attempt to improve the non-parameteric estimations via weighting does not seem to improve the selection too much.

%\par In the next step,  the non-parametric baseline hazard estimation was replaced by the exact baseline hazard that was used to generate the outcome. Results were listed in the seventh column in Table 2. When the baseline was known, the performance of sum of squared deviance residuals actually performed really close to the optimal $\lambda$. This simulation suggests that when the baseline hazard was accurately estimated, then the sum of squared deviance residuals gave really good measures for the model fit.

%\par In the last step, we replaced the non-parametric baseline hazard estimation with a parametric approach, where we assume the baseline hazard to follow an exponential distribution. We derived the estimation for baseline hazard by maximum likelihood approach conditioned on covariates: $\hat{\lambda}_{0} = \frac{\sum d_{i}}{\sum t_{i}e^{\eta_{i}}}$. The results are listed in the eighth column in Table 2. When this estimation was used for data that were generated from exponential distribution, its performance was as well as when the exact baseline was specified. When it was fitted to data that were generated from the Weibull distribution, that is when the model is misspecified, then its performance was even worse than the non-parametric approaches.

%\par Hence, getting accurate estimation of the baseline hazard is crucial to whether or not the sum of squared deviance residuals can be a good measurement for model fit.

%\begin{table}[h]
%	\small
%	\caption{Average of Selected $\lambda$ with Different Baseline Estimation for Deviance Residuals}
	%\centering
	%\begin{tabular}{lllllllll}
	%\hline
	%\\[-0.75em]
    %& Optimal & Linear Pred & KP & KP & Breslow & Breslow & Exact & Exp\\ 
    %&  & & & weighted & & weighted & & \\ \cline{2 - 9} 
	%\textbf{True Baseline: $Exponential(1)$} &&&&&&&&\\
	%n = 100, p = 10, non-zero $\beta$: 2 & 0.0461 & 0.0665 & 0.1986 & 0.1970 & 0.1908 & 0.1975 & 0.0540 & 0.0539 \\
	%n = 100, p = 100, non-zero $\beta$: 20 &0.0555 & 0.1841 & 0.2359 & 0.2201 & 0.2146 & 0.2244 & 0.0805 & 0.0985 \\
	%n = 100, p = 1000, non-zero $\beta$: 20 & 0.0268 & 0.0505 & 0.0948 & 0.0798 & 0.0867 & 0.0809 & 0.0288 & 0.0270 \\
	%\textbf{True Baseline:} $Weibull$ &&&&&&&& \\
 	%n = 100, p = 10, non-zero $\beta$: 2 & 0.0476 & 0.0710 & 0.2004 & 0.1969 & 0.1907 & 0.1907 & N/A & 0.2330 \\ 
	%\hline
%\end{tabular}
%\end{table}
